\chapter{Emotion Detection Model}

\section{Model Selection}
To model the emotional content present in phishing emails, we selected three transformer-based architectures: \acs{RoBERTa}-large \cite{liu2019roberta}, \acs{ELECTRA}-large \cite{clark2020electrapretrainingtextencoders}, and \acs{DeBERTa}-v3-large \cite{he2021debertadecodingenhancedbertdisentangled}. These models are widely adopted in \acs{NLP} classification tasks and provide strong contextual representations that are suitable for capturing subtle emotional cues in text.

Unlike the binary phishing detection task in Chapter~4, emotion detection is formulated here as a \textbf{multi-label classification} problem: a single phishing email may simultaneously express multiple emotions (e.g., fear and curiosity). This formulation better reflects realistic phishing language, where overlapping emotional manipulation strategies are common.

\section{Emotion Taxonomy Consolidation}
\label{sec:emotion_taxonomy}

The original annotation schema comprised fifteen emotional categories. Prior to model training, an inter-label co-occurrence analysis was conducted on the annotated corpus to identify clusters of emotions that consistently co-occurred and represented semantically equivalent manipulation vectors. This analysis motivated a principled consolidation from fifteen classes to eight, with the goal of increasing the per-class training signal while preserving meaningful distinctions between manipulation strategies.

Table~\ref{tab:emotion_merging} summarises the mapping from original categories to the consolidated taxonomy.

\begin{table}[htbp]
\centering
\caption{Emotion label consolidation: mapping from the original 15-class annotation schema to the 8-class training taxonomy.}
\label{tab:emotion_merging}
\begin{tabular}{lll}
\hline
\textbf{Consolidated Label} & \textbf{Original Categories} & \textbf{Rationale} \\
\hline
\texttt{positive\_arousal} & excitement, surprise, joy, desire & Co-occurrence rate 43–60\%; \\
                           &                                    & shared positive-affect manipulation vector \\
\texttt{warmth}            & admiration, gratitude, caring      & Trust-building emotional cluster \\
\texttt{threat}            & anger, fear                        & Intimidation/coercion cluster; \\
                           &                                    & 79\% of anger samples contain fear \\
\texttt{curiosity}         & curiosity                          & Distinct manipulation tactic \\
\texttt{confusion}         & confusion                          & Distinct manipulation tactic \\
\texttt{sadness}           & sadness                            & Distinct manipulation tactic \\
\texttt{relief}            & relief                             & Distinct manipulation tactic \\
\hline
\multicolumn{3}{l}{\textit{Dropped:} \texttt{neutral} (143 samples) and \texttt{unsure} (119 samples) — too rare} \\
\multicolumn{3}{l}{to learn from; \texttt{unsure} is an annotation artefact rather than a genuine emotion.} \\
\hline
\end{tabular}
\end{table}

The three merged clusters were motivated by concrete empirical and semantic criteria. The \texttt{positive\_arousal} cluster groups emotions that all serve the same attacker goal: inducing a positive affective state in the victim to lower critical thinking and prompt impulsive action. Co-occurrence rates between any two members of this cluster ranged from 43\% to 60\%, indicating that annotators overwhelmingly labelled these emotions together rather than in isolation. The \texttt{warmth} cluster consolidates the trust-building manipulation tactics that phishing attackers use to establish rapport before delivering a malicious request; admiration, gratitude, and caring rarely appeared without one another in the corpus. The \texttt{threat} cluster is motivated by the strongest empirical signal: 79\% of samples labelled with anger also carried a fear label, and both emotions serve the same intimidation and coercion function within phishing messages.

Four categories were retained as independent classes because each represents a genuinely distinct manipulation tactic with a unique psychological profile and no significant co-occurrence with other classes: \texttt{curiosity} (exploiting information-seeking behaviour), \texttt{confusion} (disorienting the victim to reduce scrutiny), \texttt{sadness} (guilt-induction and empathy exploitation), and \texttt{relief} (offering resolution of a fabricated threat to prompt compliance).

The \texttt{neutral} and \texttt{unsure} labels were dropped entirely. Their combined sample count (262 instances) was insufficient to provide reliable training signal, and \texttt{unsure} is an annotation artefact that reflects annotator uncertainty rather than a phishing manipulation strategy.

This consolidation yields a final training taxonomy of \textbf{eight classes}. Each class benefits from two to four times the training signal of its constituent original categories, which is expected to improve model calibration and reduce the severity of the class imbalance problem relative to the original schema.

The label merging was applied programmatically prior to training via a dedicated preprocessing script (\texttt{merge\_labels.py}), which reads the raw annotated corpus and writes the consolidated version to \texttt{combined\_emails\_dataset\_merged.csv}. All training scripts operate on this merged file.

\section{Dataset Preparation}
The starting point was the combined email dataset used throughout this thesis. For emotion modelling, only emails labelled as \texttt{phishing} were retained, since the objective is to characterise emotional patterns specifically within malicious messages.

Rows without emotion annotation were removed, and the remaining emotion labels were parsed from a delimiter-based format (e.g., \texttt{curiosity\#threat\#warmth}) into sets of class tags reflecting the consolidated taxonomy described in Section~\ref{sec:emotion_taxonomy}. A global emotion vocabulary of eight classes was built from the training corpus, and each email was encoded as a multi-hot vector, where each dimension indicates the presence (1) or absence (0) of a given consolidated emotion.

To preserve label distribution during splitting, a stratification proxy was used: for each sample, the index of its dominant label (\texttt{argmax} over the multi-hot vector) served as the stratification target. Data was split into test (20\%) and train (80\%), followed by a validation split (10\% of train), resulting in approximately 72\%/8\%/20\% train/validation/test partitions, with a fixed random seed for reproducibility.

\section{Model Implementation}
Model development used the Hugging Face Transformers ecosystem \cite{wolf2020transformers} for tokenisation, model fine-tuning, and checkpoint management, together with scikit-learn \cite{pedregosa2011scikit} for metric computation.

For each model (\acs{RoBERTa}-large, \acs{ELECTRA}-large, \acs{DeBERTa}-v3-large), the following pipeline was applied:

\begin{itemize}
	\item \textbf{Tokenisation:} Emails were tokenised with each model's pretrained tokeniser, padded to fixed length, and truncated at 512 tokens.

	\item \textbf{Multi-label Objective:} The classifier head was configured with \texttt{problem\_type=multi\_label\_classification}. During training, logits were optimised using \texttt{BCEWithLogitsLoss}, and probabilities were obtained at evaluation time using the sigmoid function.

	\item \textbf{Class Imbalance Handling:} Because some consolidated classes are less frequent than others, per-class positive weights were computed from the training split as the negative-to-positive ratio. To avoid excessive penalisation of rare classes, a square-root dampening strategy was used and values were clipped to a maximum of 5.0. These \texttt{pos\_weight} values were injected into \texttt{BCEWithLogitsLoss} through a custom weighted trainer.

	\item \textbf{Model Fine-tuning:} All models were fine-tuned for up to 10 epochs using \acs{AdamW} \cite{loshchilov2019decoupled}, cosine learning-rate scheduling, learning rate $2 \times 10^{-5}$, warmup ratio 0.15, and weight decay 0.01. Early stopping (patience = 3) monitored validation macro-F1, with best-checkpoint restoration enabled.

	\item \textbf{Computational Optimisations:} BF16 mixed precision \cite{micikevicius2018mixed}, gradient checkpointing, and gradient clipping (max norm 1.0) were used to improve training efficiency and stability on \acs{GPU} hardware.

	\item \textbf{Threshold Tuning:} Since multi-label outputs depend strongly on decision thresholds, threshold optimisation was performed on the validation set after training. Two strategies were evaluated: (i) one global threshold shared by all classes, and (ii) per-class thresholds. Candidate thresholds in the interval [0.10, 0.80] with 0.05 steps were tested to maximise macro-F1.

	\item \textbf{Evaluation:} Performance was reported using exact-match accuracy, macro precision, macro recall, and macro F1, along with per-class classification reports on the held-out test set.
\end{itemize}

\section{Evaluation and Results}

After training, each model was evaluated on the test set using both thresholding strategies selected from validation: global threshold and per-class thresholds. This comparison assesses whether class-specific calibration improves the balance between precision and recall across the eight consolidated emotion classes.

\subsection{RoBERTa-large Results}
Table~\ref{tab:roberta_emotion_results} summarises the test performance of the \acs{RoBERTa}-large model under both thresholding strategies.

\begin{table}[htbp]
\centering
\small
\caption{Test set results for \acs{RoBERTa}-large using global and per-class thresholds.}
\label{tab:roberta_emotion_results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccc}
\hline
\textbf{Threshold Strategy} & \textbf{Exact Match} & \textbf{Macro F1} & \textbf{Macro Precision} & \textbf{Macro Recall} \\
\hline
Global threshold (best on val)     & 0.2918 & 0.6869 & 0.6823 & 0.7091 \\
Per-class thresholds (best on val) & 0.2816 & 0.7041 & 0.6793 & 0.7477 \\
\hline
\end{tabular}
}
\end{table}

\subsection{ELECTRA-large Results}
Table~\ref{tab:electra_emotion_results} reports the corresponding results for \acs{ELECTRA}-large.

\begin{table}[htbp]
\centering
\small
\caption{Test set results for \acs{ELECTRA}-large using global and per-class thresholds.}
\label{tab:electra_emotion_results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccc}
\hline
\textbf{Threshold Strategy} & \textbf{Exact Match} & \textbf{Macro F1} & \textbf{Macro Precision} & \textbf{Macro Recall} \\
\hline
Global threshold     & 0.3326 & 0.6967 & 0.7168 & 0.6911 \\
Per-class thresholds & 0.2775 & 0.7036 & 0.6890 & 0.7407 \\
\hline
\end{tabular}
}
\end{table}

\subsection{DeBERTa-v3-large Results}
Table~\ref{tab:deberta_emotion_results} presents the test results for \acs{DeBERTa}-v3-large.

\begin{table}[htbp]
\centering
\small
\caption{Test set results for \acs{DeBERTa}-v3-large using global and per-class thresholds.}
\label{tab:deberta_emotion_results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccc}
\hline
\textbf{Threshold Strategy} & \textbf{Exact Match} & \textbf{Macro F1} & \textbf{Macro Precision} & \textbf{Macro Recall} \\
\hline
Global threshold (best on val)     & TODO & TODO & TODO & TODO \\
Per-class thresholds (best on val) & TODO & TODO & TODO & TODO \\
\hline
\end{tabular}
}
\end{table}

\subsection{Comparative Analysis}
Table~\ref{tab:emotion_model_comparison} provides a cross-model summary using the best threshold strategy for each architecture. The label consolidation described in Section~\ref{sec:emotion_taxonomy} ensures that all models are trained and evaluated against the same eight semantically coherent classes, making cross-model comparison meaningful and directly comparable.

\begin{table}[htbp]
\centering
\small
\caption{Cross-model comparison on the test set (best threshold strategy per model).}
\label{tab:emotion_model_comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lccccc}
\hline
\textbf{Model} & \textbf{Best Strategy} & \textbf{Exact Match} & \textbf{Macro F1} & \textbf{Macro Precision} & \textbf{Macro Recall} \\
\hline
\acs{RoBERTa}-large   & Per-class thresholds & 0.2816 & 0.7041 & 0.6793 & 0.7477 \\
\acs{ELECTRA}-large   & Global threshold     & 0.3326 & 0.6967 & 0.7168 & 0.6911 \\
\acs{DeBERTa}-v3-large & TODO                & TODO   & TODO   & TODO   & TODO   \\
\hline
\end{tabular}
}
\end{table}