\chapter{Phishing Detection Model}
\label{chapter:model}

\section{Model Selection}
The selection of appropriate architectures for phishing email classification requires balancing representational capacity, training stability, and the specific linguistic characteristics of malicious email text. Based on a review of recent advances in \ac{NLP} and the findings surveyed in Chapter~\ref{chapter:related_work}, two transformer-based models were selected: \acs{RoBERTa}-large \cite{liu2019roberta} and \acs{ELECTRA}-large \cite{clark2020electra}.

\textbf{\acs{RoBERTa}-large} is a robustly optimised variant of \acs{BERT} trained on substantially larger data (160\,GB of text) with dynamic masking, longer sequences, and larger batch sizes. Its bidirectional self-attention mechanism allows each token to contextualise against all other tokens simultaneously, which is particularly valuable for phishing detection because manipulation tactics in malicious emails frequently span the entire message: urgency cues are planted in the subject line, credibility is built in opening paragraphs, and the malicious request arrives near the end. Capturing these long-range dependencies without vanishing gradient issues is a core strength of the RoBERTa architecture. The large variant (355\,M parameters) provides richer representational capacity than base-size alternatives, at the cost of higher memory and computational requirements, a trade-off justified here by the complexity and nuance of the detection task.

\textbf{\acs{ELECTRA}-large} departs from the masked language modelling pretraining paradigm of BERT/RoBERTa by training a discriminator network to distinguish real tokens from plausible replacements generated by a small generator model. This replaced-token-detection objective exposes the model to all input positions during pretraining, rather than only the 15\% of tokens that are masked in the BERT objective, making ELECTRA more sample-efficient and producing representations that are particularly sensitive to lexical and semantic coherence. These characteristics are directly useful for detecting subtle linguistic anomalies in phishing text: an email that is stylistically inconsistent, employs incongruous formality levels, or contains phrases that are plausible in isolation but unusual in combination will exhibit low discriminator confidence under the ELECTRA objective, providing a different signal from the contextual uncertainty captured by RoBERTa. The large discriminator (335\,M parameters) has demonstrated competitive or superior performance to RoBERTa-large on several text classification benchmarks \cite{clark2020electra}, making it a strong complementary model for ensemble construction.

The decision to use both architectures jointly was motivated by their complementary inductive biases: RoBERTa's masked language modelling produces robust long-range contextual representations, while ELECTRA's discriminative pretraining yields fine-grained sensitivity to local token plausibility. Averaging their softmax probabilities in an ensemble means that disagreements on borderline cases are naturally surfaced by the \textit{maybe spam} tier of the three-class inference scheme, rather than forcing a premature hard binary decision. Both models have demonstrated state-of-the-art performance in text classification and semantic similarity tasks \cite{liu2019roberta, sanh2019distilbert}, and their large-scale pretrained representations provide a strong initialisation for fine-tuning on the relatively modest phishing corpus assembled in Chapter~\ref{chapter:dataset}.

\section{Dataset Information \& Preparation}
The dataset used for training and evaluation consists of a combination of the original 10,000 phishing email samples generated in Chapter~\ref{chapter:dataset} and an additional 10,000 non-phishing email samples produced using the same LLM-based generation methodology, but without emotional category conditioning. The resulting corpus of 20,000 samples is balanced at the class level, with exactly 10,000 examples per class.

Before splitting, the dataset underwent a lightweight preprocessing step. The \texttt{text} column was formed by concatenating available header fields (subject, sender) with the email body, separated by newlines, to provide the model with all textual signals available in a real email client. Labels were mapped to binary integers (\texttt{non-phishing}$\to$0, \texttt{phishing}$\to$1) and rows with missing labels were discarded. Emails containing only whitespace after extraction were also removed.

Stratified splits were produced using \texttt{train\_test\_split} from scikit-learn \cite{pedregosa2011scikit} with a fixed random seed of 42 to ensure reproducibility. Twenty percent of the data was held out as the test set. A further ten percent of the remaining training data was reserved as a validation set, used exclusively for early stopping and threshold selection. This yields approximate proportions of 72\% training (14,400 samples), 8\% validation (1,600 samples), and 20\% test (4,000 samples). Stratification ensures that the 50/50 class balance is preserved across all three splits.

Class weights were computed on the training partition using \texttt{compute\_class\_weight} with the \texttt{balanced} strategy from scikit-learn. Although the overall corpus is balanced, the stratified sub-sampling can introduce minor deviations in individual partitions, so explicit class weighting serves as a safeguard and maintains consistency with the weighting strategy applied to the emotion detection models in Chapter~\ref{chapter:emotion_model}. The resulting weights were incorporated into a custom \texttt{WeightedTrainer} that replaces the standard cross-entropy loss with a weighted variant.

\section{Model Implementation}
The implementation leveraged the Hugging Face Transformers library \cite{wolf2020transformers} for model loading, tokenization, and training, combined with scikit-learn \cite{pedregosa2011scikit} for evaluation metrics and class weight computation. For each model (\acs{RoBERTa}-large, \acs{ELECTRA}-large), the following steps were performed:

\begin{itemize}
	\item \textbf{Tokenization:} Emails were tokenised using the respective pretrained tokeniser, with padding to the maximum length and truncation at 512 tokens. Padding was handled dynamically at collation time using \texttt{DataCollatorWithPadding} to avoid inflating sequences to the global maximum during the preprocessing step, reducing memory overhead during training. For \acs{ELECTRA}-large, the \texttt{use\_fast=False} flag was set to avoid a tokeniser conversion issue present in some Transformers versions.

	\item \textbf{Class Imbalance Handling:} To address any residual class imbalance, balanced class weights were computed using scikit-learn's \texttt{compute\_class\_weight} function. These weights were incorporated into the loss function through a custom weighted trainer that replaced the standard cross-entropy loss with a weighted variant, penalizing misclassifications of the minority class more heavily. At each forward pass, the trainer pops the \texttt{labels} tensor from the inputs, computes logits from the model, and evaluates the loss with the weight tensor moved to the same device as the logits, ensuring device consistency during mixed-precision training.

	\item \textbf{Model Fine-tuning:} Each pretrained model was fine-tuned for binary sequence classification using the \ac{AdamW} optimizer \cite{loshchilov2019decoupled} with a cosine learning rate scheduler. Training was performed for up to 10 epochs with a linear warmup over the first 10\% of training steps, weight decay of 0.01, and gradient clipping with a maximum norm of 1.0. The learning rate was set to $3 \times 10^{-5}$ for \acs{RoBERTa}-large and $2 \times 10^{-5}$ for \acs{ELECTRA}-large, reflecting the greater sensitivity of the ELECTRA discriminator to higher learning rates. Label smoothing with a factor of 0.05 was applied to improve probability calibration, which is important for the threshold-based three-tier inference scheme. Early stopping with a patience of 3 epochs was used, monitoring the F1-score on the validation set, and the best model checkpoint was restored at the end of training.

	\item \textbf{Computational Optimizations:} Mixed precision training using \ac{BF16} \cite{micikevicius2018mixed} was enabled to accelerate computation and reduce memory usage on compatible \ac{GPU} hardware. Unlike the emotion detection models described in Chapter~\ref{chapter:emotion_model}, gradient checkpointing was disabled for these binary classifiers, as sufficient \ac{GPU} memory was available and disabling it yields faster training throughput. The fused AdamW implementation (\texttt{adamw\_torch\_fused}) was used to reduce optimizer overhead, and data loading was parallelised with four worker processes and pinned memory for improved throughput.

	\item \textbf{Evaluation:} Model performance was monitored during training using accuracy, precision, recall, and F1-score on the validation set. The best model checkpoint was selected based on the highest validation F1-score. At inference time, in addition to using the standard decision threshold of 0.5, a custom threshold of 0.35 was evaluated. This lower threshold was chosen to increase recall for phishing detection, since in a security context, failing to detect a phishing email (false negative) carries greater risk than a false positive. Lowering the threshold shifts the operating point toward higher recall at a modest cost to precision, which is the preferred trade-off for a security-oriented classifier.
\end{itemize}

\section{Evaluation and Results}

After training, each model was evaluated on the held-out test set. Two decision thresholds were compared: the standard threshold of 0.5, where the class with the highest softmax probability is selected, and a custom threshold of 0.35, which classifies an email as phishing whenever the predicted phishing probability exceeds 0.35. This lower threshold aims to improve recall at a modest cost to precision, reflecting the asymmetric cost of errors in phishing detection.

\subsection{ELECTRA-large Results}

Table~\ref{tab:electra_results} presents the test set performance of the \acs{ELECTRA}-large model at both decision thresholds.

\begin{table}[htbp]
\centering
\caption{Test set results for \acs{ELECTRA}-large at standard and custom thresholds.}
\label{tab:electra_results}
\begin{tabular}{lcccc}
\hline
\textbf{Threshold} & \textbf{Accuracy} & \textbf{F1} & \textbf{Precision} & \textbf{Recall} \\
\hline
0.5 (Standard) & 0.9256 & 0.9051 & 0.9230 & 0.8880 \\
0.35 (Custom)  & 0.9256 & 0.9055 & 0.9195 & 0.8920 \\
\hline
\end{tabular}
\end{table}

The \acs{ELECTRA}-large model achieved an F1-score of 0.9051 at the standard threshold, with high precision (0.9230) but comparatively lower recall (0.8880). Lowering the threshold to 0.35 yielded a marginal improvement across all metrics, increasing the F1-score to 0.9055 and recall to 0.8920, while precision decreased only slightly to 0.9195. This confirms that the custom threshold provides a better trade-off for phishing detection, capturing more true positives with minimal impact on precision.

\subsection{RoBERTa-large Results}
Table~\ref{tab:roberta_results} presents the test set performance of the \acs{RoBERTa}-large model at both decision thresholds.

\begin{table}[htbp]
\centering
\caption{Test set results for \acs{RoBERTa}-large at standard and custom thresholds.}
\label{tab:roberta_results}
\begin{tabular}{lcccc}
\hline
\textbf{Threshold} & \textbf{Accuracy} & \textbf{F1} & \textbf{Precision} & \textbf{Recall} \\
\hline
0.5 (Standard) & 0.9352 & 0.9169 & 0.9410 & 0.8940 \\
0.35 (Custom)  & 0.9336 & 0.9150 & 0.9371 & 0.8940 \\
\hline
\end{tabular}
\end{table}

The \acs{RoBERTa}-large model achieved an F1-score of 0.9169 at the standard threshold, with very high precision (0.9410) and recall (0.8940). Lowering the threshold to 0.35 resulted in a slight decrease in accuracy and F1-score, with precision dropping to 0.9371 while recall remained unchanged at 0.8940. This suggests that for \acs{RoBERTa}-large, the standard threshold already provides a strong balance between precision and recall, and the custom threshold does not offer a significant improvement in this case.

\subsection{Baseline Comparison}
\label{sec:baseline_comparison}

To contextualise the performance of the fine-tuned transformer models, two classical baselines were evaluated under the same experimental conditions: identical data splits (random seed 42), identical stratification strategy, and identical threshold settings.

\textbf{TF-IDF + Logistic Regression} represents the standard strong baseline for text classification. Input emails were tokenised into unigrams and bigrams, weighted by sublinear TF-IDF, and classified using an $\ell_2$-regularised logistic regression with balanced class weights, mirroring the class-weight strategy applied during transformer fine-tuning.

\textbf{TF-IDF + Multinomial Naive Bayes} provides a historically relevant reference point, reflecting the generation of spam filters from which phishing detection evolved \cite{metsis2006spam}. Both classifiers used a vocabulary of up to 100,000 features with a minimum document frequency of two.

Table~\ref{tab:baseline_comparison} reports all four models at both evaluated thresholds.

\begin{table}[htbp]
\centering
\caption{Phishing detection results on the held-out test set for classical baselines and fine-tuned transformer models.}
\label{tab:baseline_comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llcccc}
\hline
\textbf{Model} & \textbf{Threshold} & \textbf{Accuracy} & \textbf{F1} & \textbf{Precision} & \textbf{Recall} \\
\hline
\multicolumn{6}{l}{\textit{Classical baselines}} \\
TF-IDF + Multinomial NB     & 0.50 (Standard) & 0.8000 & 0.7459 & 0.7583 & 0.7340 \\
TF-IDF + Logistic Regression & 0.50 (Standard) & 0.8856 & 0.8560 & 0.8621 & 0.8500 \\
\hline
\multicolumn{6}{l}{\textit{Fine-tuned transformer models}} \\
\acs{ELECTRA}-large          & 0.35 (Custom)   & 0.9256 & 0.9055 & 0.9195 & 0.8920 \\
\acs{RoBERTa}-large          & 0.35 (Custom)   & 0.9336 & 0.9150 & 0.9371 & 0.8940 \\
\hline
\end{tabular}%
}
\end{table}

The TF-IDF + Logistic Regression baseline achieved an F1-score of 0.8560, demonstrating that lexical features alone carry substantial signal for phishing detection. However, this represents a gap of approximately 6 percentage points in F1 relative to \acs{RoBERTa}-large (0.9150), confirming that contextual, deep representations learned during pre-training provide meaningful improvements beyond surface-level term statistics. Multinomial Naive Bayes performs considerably worse (F1 = 0.7459), consistent with its inability to model feature correlations and its poor calibration on long, variable-length email texts. The ROC-AUC of the logistic regression baseline (0.948) is noteworthy --- it approaches transformer-level discriminative ability at the ranking level --- yet this does not translate into calibrated decisions at any single threshold, which explains the F1 gap. Taken together, these results establish that the transformer ensemble provides a clear and quantifiable improvement over classical approaches, while the logistic regression baseline sets a competitive non-trivial floor that any proposed method should be expected to surpass.

\subsection{Comparative Analysis}
\label{sec:evaluation}
Table~\ref{tab:phishing_model_comparison} summarizes both models using the threshold configuration selected for deployment.

\begin{table}[htbp]
\centering
\caption{Cross-model comparison on the test set (selected threshold per model).}
\label{tab:phishing_model_comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llcccc}
\hline
\textbf{Model} & \textbf{Selected Threshold} & \textbf{Accuracy} & \textbf{F1} & \textbf{Precision} & \textbf{Recall} \\
\hline
\multicolumn{6}{l}{\textit{Classical baselines}} \\
TF-IDF + Multinomial NB      & 0.50 (Standard) & 0.8000 & 0.7459 & 0.7583 & 0.7340 \\
TF-IDF + Logistic Regression & 0.50 (Standard) & 0.8856 & 0.8560 & 0.8621 & 0.8500 \\
\hline
\multicolumn{6}{l}{\textit{Fine-tuned transformer models}} \\
\acs{ELECTRA}-large           & 0.35 (Custom)   & 0.9256 & 0.9055 & 0.9230 & 0.8880 \\
\acs{RoBERTa}-large           & 0.35 (Custom)   & 0.9336 & 0.9150 & 0.9371 & 0.8940 \\
\hline
\end{tabular}%
}
\end{table}

Based on the similar performance of both models, the final selection was a combination of \acs{ELECTRA}-large and \acs{RoBERTa}-large, leveraging the strengths of both architectures. The ensemble approach averages the softmax probabilities of both models and applies a tiered classification scheme: emails with an averaged probability below the ensemble threshold are classified as \textit{ham}; emails between the ensemble threshold and 0.50 are flagged as \textit{maybe spam}, surfacing borderline cases for user review without a hard binary decision; and emails at or above 0.50 are classified as \textit{spam}. This three-tier scheme reflects the asymmetric risk of false negatives in a security context while avoiding the false-positive burden of a purely low-threshold binary classifier.