\chapter{Introduction}%
\label{chapter:introduction}

The presence of the internet in our daily lives has become increasingly significant, transforming the way we communicate, access information, and conduct business. Over the past few decades, digital communication has evolved from a luxury to a necessity, fundamentally reshaping societal interactions and organizational operations. Among the various communication channels available today, email remains one of the most ubiquitous and reliable methods of exchanging information across the globe.

Since its inception in the early 1970s, email has evolved into an essential tool for both personal and professional communication, allowing users to send and receive messages quickly and efficiently across vast distances at minimal cost. The convenience, accessibility, and widespread adoption of email have made it an indispensable component of modern digital infrastructure. Today, email is used for everything from casual personal correspondence to business transactions, legal communications, and authentication processes for online services.

As such, it is now used across the world, with billions of users relying on email for various purposes. Statistics from 2023 show that around 4.37 billion people use email globally, with 347 billion emails sent and received daily\cite{statista_email}. This scale and the nature of information transmitted through email make it a prime target for malicious actors who seek to exploit vulnerabilities in email systems for their own gain. Email security has consequently emerged as a concern for individuals, organizations, and governments alike, as the consequences of a successful email attack can be severe and far-reaching, ranging from data breaches and identity theft to financial losses, reputational damage, and even national security threats.

The landscape of email-based threats is diverse and constantly evolving. Among these threats, phishing stands out as one of the most common, effective, and persistent methods used by cybercriminals to deceive users into divulging sensitive information or downloading malware. Phishing attacks typically involve sending fraudulent emails that appear to be from a legitimate source, such as a bank, government agency, popular social media platform, or trusted colleague, in order to lure recipients into clicking on a malicious link, opening an infected attachment, or providing confidential information such as passwords, credit card numbers, or personal identification details.

The sophistication of phishing attacks has increased dramatically over the years. Early phishing attempts were often easily identifiable due to poor grammar, obvious formatting errors, and generic greetings. However, modern phishing campaigns employ advanced social engineering techniques, psychological manipulation, and increasingly sophisticated technical methods to evade detection. Attackers now leverage artificial intelligence to craft convincing messages, personalize content based on publicly available information about their targets, and create convincing replicas of legitimate websites. The rise of spear phishing, targeted attacks aimed at specific individuals or organizations, and whaling, attacks targeting high-profile executives, demonstrates the evolution and specialization of these threats.

To combat the growing threat of phishing attacks, various email security measures have been developed and deployed over the years. These include traditional spam filters that use rule-based systems and blacklists, antivirus software that scans attachments for known malware signatures, authentication protocols such as \ac{SPF}, \ac{DKIM}, and \ac{DMARC} that verify sender legitimacy, and multi-factor authentication systems that add additional layers of security. However, these measures are not foolproof, and cybercriminals are constantly finding new ways to bypass them. The arms race between security professionals and attackers continues to intensify, with each side developing increasingly sophisticated techniques.

One promising but underexplored avenue in the fight against phishing involves analyzing the emotional content and psychological manipulation tactics employed in malicious emails. Phishing attacks often exploit human emotions such as fear, urgency, curiosity, or greed to prompt hasty actions without careful consideration. 

This work aims to explore the defense angle against phishing attacks through the development of a machine learning-based emotion detection system that can identify and flag potentially malicious emails based on their emotional content and psychological manipulation tactics. By analyzing the emotional tone, sentiment, and affective characteristics of emails, we can gain deeper insights into the intent behind them and identify patterns that may indicate a phishing attempt. This approach represents several contributions, such as:

\begin{itemize}
    \item \textbf{Public dataset:} The release of a publicly available phishing corpus annotated with fine-grained emotional labels in both European Portuguese (PT-PT) and English (EN), intended to support reproducible research on affective signals in malicious emails.

    \item \textbf{Consolidated taxonomy:} A principled consolidation of the original emotional taxonomy informed by co-occurrence analysis of annotations, producing a smaller set of semantically coherent labels that reduces ambiguity and improves model learnability.

    \item \textbf{Model ensemble:} An ensemble combining \ac{RoBERTa}-large and \ac{ELECTRA}-large models together with a three-tier inference scheme (\textit{ham} / \textit{maybe spam} / \textit{spam}), which averages model probabilities to prioritise recall while controlling false positives.

    \item \textbf{End-to-end application:} A deployed end-to-end system comprising a FastAPI backend, a user-facing web interface for uploading emails, and a Manifest V3 Chrome extension that enables in-browser scanning and inspection of email content.
\end{itemize}

\section{Thesis Objectives}
The primary objective of this thesis is to develop a robust and effective machine learning-based system for detecting phishing emails by analyzing their emotional content and psychological manipulation tactics. This work is organised around five goals. First, it aims to create a comprehensive dataset of phishing emails that includes a diverse range of emotional categories and manipulation tactics, combining real-world examples, synthetic data generated through \acs{LLM} prompting, and rigorous emotional annotation. Second, it seeks to develop and fine-tune transformer-based models for both binary phishing detection and multi-label emotion classification, with appropriate handling of class imbalance and careful hyperparameter selection. Third, it focuses on evaluating model performance on a held-out test set through threshold comparison, precision, recall, F1-scores, and error analysis. Fourth, it targets the integration of the trained models into a user-friendly web application and companion browser extension for practical email analysis. Finally, it addresses ethical concerns related to data privacy, bias, and responsible deployment in real-world settings.

\section{Contributions}
This thesis makes five key contributions to the fields of email security and natural language processing. It introduces a comprehensive phishing dataset annotated with emotional categories and psychological manipulation tactics, which can support future research. It develops and fine-tunes transformer-based models for both binary phishing detection and multi-label emotion classification, demonstrating the usefulness of these architectures in this setting. It provides a detailed evaluation of performance across decision thresholds and emotional categories, offering insight into the models' strengths and limitations. It integrates the trained models into a user-friendly web application and companion browser extension, showing the practical value of the research. Finally, it includes a discussion of ethical considerations related to data privacy, model bias, and responsible deployment, contributing to the broader discourse on ethical AI in cybersecurity.

\section{Thesis Structure}
The thesis is structured into seven chapters. Chapter~\ref{chapter:introduction} introduces the research problem, objectives, investigation questions, contributions, and thesis structure. Chapter~\ref{chapter:related_work} reviews the existing literature on phishing detection, emotional analysis in cybersecurity, and transformer-based models. Chapter~\ref{chapter:dataset} describes the dataset creation process, including data collection, annotation, and emotional taxonomy consolidation. Chapter~\ref{chapter:model} details the model selection, implementation, and training procedures for both phishing detection and emotion classification. Chapter~\ref{chapter:webapp} describes the development and deployment of the web application and browser extension. Finally, Chapter~\ref{chapter:results} and Chapter~\ref{chapter:conclusion} summarize the findings, discusses limitations, and outlines future work.